Neutral models of \textit{de novo} gene emergence suggest that gene evolution has a preferred trajectory

### 2023-03-08 Manuscripts/M1/deNovoRates_old.tex
renamed:    Manuscripts/Response_to_Reviewers_Ravi2023MBE.tex -> Manuscripts/M1/Response_to_Reviewers_Ravi2023MBE.tex 	renamed:    Manuscripts/SM_deNovoRates.tex -> Manuscripts/M1/SM_deNovoRates.tex 	renamed:    Manuscripts/deNovoRates.tex -> Manuscripts/M1/deNovoRates.tex 	renamed:    Manuscripts/deNovoRates_old.tex -> Manuscripts/M1/deNovoRates_old.tex 	renamed:    Manuscripts/deNovoStudiesRNAseqInfo.csv -> Manuscripts/M1/deNovoStudiesRNAseqInfo.csv 	renamed:    Manuscripts/mybst.bst -> Manuscripts/M1/mybst.bst 	renamed:    Manuscripts/refs.bib -> Manuscripts/M1/refs.bib 	new file:   Manuscr

In this study we focus on the evolution of genes that encode new proteins. New protein coding genes frequently arise via duplication of existing protein coding genes. These duplicated genes then genetically and functionally diverge from their parent genes by accumulating mutations \citep{Long2003,IAD}. Recently it has been shown that new protein coding genes can arise independent of gene duplication, from genomic sequences that did not previously encode any protein. This phenomenon is called \textit{de novo} gene emergence \citep{EBB-F1000,vanOss}. \textit{De novo} protein coding genes can emerge from intergenic sequences, non-coding RNA genes, introns, and even regions that partially overlap with existing protein coding genes \citep{vanOss}. To express a protein, a genomic sequence should be transcribed as well as translated. Therefore, the gene needs sequence features that enable both these processes.

@@

We made our analysis of gene loss more stringent by comparing the probabilities of two evolutionary scenarios, both of which begin with an intact proto-gene. In the first scenario, the ORF remains intact whereas the transcription is lost. The probability of this scenario is described by: 
$P_\textit{ORF-stay \vline\ RNA} \times P_\textit{RNA-loss \vline\ ORF}/P_\textit{ORF}$. The probability of the second scenario, where transcription stays intact while ORF is lost, is: $P_\textit{ORF-loss} \times P_\textit{RNA-stay}/P_\textit{RNA}$. We find that the first scenario is more plausible for ORFs that have fewer than 34 codons, which is in agreement with our previous analysis (\autoref{lossprob}). Also, the probability values of scenario one and two, obtained from this stringent analysis, were not significantly different from the transcription ($P_\textit{RNA-loss}$) and ORF loss ($P_\textit{ORF-loss}$) probabilities, respectively.

@@

$$P_\textit{stop-gain \vline\ RNA} = \frac{P_\textit{stop-gain}\times(1- P_\textit{stop} - P_\textit{poly-A})}{1- P_\textit{stop}}$$

>>>>

In this study we focus on the evolution of genes that encode new proteins. New protein coding genes frequently arise via duplication of existing protein coding genes. These duplicated genes then genetically and functionally diverge from their parent genes by accumulating mutations \citep{Long2003,IAD}. Recently it has been shown that new protein coding genes can arise independent of gene duplication, from genomic sequences that did not previously encode any protein. This phenomenon is called \textit{de novo} gene emergence \citep{Tautz2011,Zhao2014,EBB-F1000,Vakirlis2017,vanOss}. \textit{De novo} protein coding genes can emerge from intergenic sequences, non-coding RNA genes, introns, and even regions that partially overlap with existing protein coding genes \citep{Tautz2011,Zhao2014,Vakirlis2017,vanOss,Prabh2019}. To express a protein, a genomic sequence should be transcribed as well as translated. Therefore, the gene needs sequence features that enable both these processes.

@@

We made our analysis of gene loss more stringent by comparing the probabilities of two evolutionary scenarios, both of which begin with an intact proto-gene. In the first scenario, the ORF remains intact whereas the transcription is lost. The probability of this scenario ($P_\textit{onlyORF-loss}$) is described by:

\begin{equation}
P_\textit{onlyRNA-loss} = \frac{P_\textit{ORF-stay \vline\ RNA}}{P_\textit{ORF \vline \ RNA}} \times P_\textit{RNA-loss \vline\ ORF}
\end{equation}

We note that the probability that an ORF is not lost due to mutations ($P_\textit{ORF-stay \vline \ RNA}$), includes the probability that the ORF already exists ($P_\textit{ORF \vline \ RNA}$). To describe the conditional probability of ORF remaining undisrupted, given that it already exists, we divide $P_\textit{ORF-stay \vline \ RNA}$ by $P_\textit{ORF \vline \ RNA}$.

In the second scenario, transcription stays intact while ORF is lost. The probability of this scenario is:

\vspace{-1em}
\begin{equation}
P_\textit{onlyORF-loss} = \frac{P_\textit{RNA-stay \vline\ ORF}}{P_\textit{RNA \vline \ ORF}} \times P_\textit{ORF-loss \vline\ RNA}
\end{equation}

We find that the second scenario (only ORF loss) is more plausible for all the analysed ORFs lengths. Furthermore, the ratio of $P_\textit{onlyORF-loss}$ and $P_\textit{onlyRNA-loss}$ has nearly identical values and trend across different ORF lengths (data not shown), as the ratio of $P_\textit{ORF-loss}$ and $P_\textit{RNA-loss}$ shown in \autoref{lossprob}.

@@

$$P_\textit{stop-gain \vline\ RNA} = \frac{P_\textit{stop-gain}\times(1- P_\textit{stop} + P_\textit{poly-A})}{1- P_\textit{stop}}$$
